D3064R0: OOTA Execution is Provably Vacuous

### 2023-12-01 ootavacuous.tex
Self-review plus publicity plan
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

	This paper goes further, adding temporal constraints from
	long-standing laws of physics, thereby providing an informal
	proof that OOTA cannot occur in correctly constructed C++
	programs running on real-world systems.
	In other words, a correct C++ implementation necessarily
	preserves the semantic dependencies that prevent OOTA
	cycles from forming.
\end{abstract}

@@

Prior work defines ``OOTA cycle'' by example, without a precise
definition.
Sometimes ``causal cycle'' is used as if it was a definition, but
without a clear decision process for what does and does not
constitute a causal cycle..
P0442R0 defines an OOTA cycle as a fixed-point computation that is
sensitive to perturbations, which makes perfect sense to that paper's
authors, but has left others unsatisfied.

We could follow this long-established tradition of proposing
solutions to the OOTA problem without a definition of this problem.
However, we (perhaps foolishly) felt the need to provide a clear
definition.

An OOTA cycle (also known as a causal cycle) is a cycle in the
memory-model sense involving the load-buffering (LB) pattern.
Within each thread in the cycle, the value stored depends on
the value loaded.
This dependency might be a data dependency
(see Listing~\ref{lst:OOTA Cycle})
or it might be a control dependency
(see Listing~\ref{lst:OOTA Cycle Based on Conditionals}).\footnote{
	Sufficiently motivated readers should have no problem constructing
	examples involving address dependencies.}
There must be at least two distinct values such that if a given load
returns a given value, that value forces the cycle to execute so as to
cause the corresponding store's value to be equal to that given value, and
for that store to occur early enough for the load to return that value.

@@

@@

@@

We are grateful to David Goldblatt for his careful review of an early
draft of this paper.
Nonetheless, all errors and omissions in this paper are the sole property
of the authors.

>>>>

	This paper goes further, adding temporal constraints from
	long-standing laws of physics and constraints on hardware
	architecture and design, thereby providing an informal
	proof that OOTA cannot occur in correctly constructed C++
	programs running on real-world systems.

	The result is that a correct C++ implementation necessarily
	preserves the semantic dependencies that prevent OOTA
	cycles from forming.
\end{abstract}

@@

Prior work defines ``OOTA cycle'' by example, without a precise
definition.
Sometimes ``causal cycle'' is used as if it was a definition, but
without a clear decision process for what does and does not
constitute a causal cycle..
P0442R0 defines an OOTA cycle as a fixed-point computation that is
destroyed by perturbations, which makes perfect sense to that paper's
authors, but has left others unsatisfied.

We could follow this long-established tradition of proposing
solutions to the OOTA problem without a definition of this problem.
However, we (perhaps foolishly) felt the need to provide a clear
definition, which is a more detailed version of the wording in 33.5.4p8
(\co{[atomics.order]}):

\begin{quote}
	Implementations should ensure that no “out-of-thin-air” values
	are computed that circularly depend on their own computation.
\end{quote}

We therefore define an OOTA cycle (also known as a causal cycle) as a
cycle in the memory-model sense involving the load-buffering (LB) pattern.
Within each thread in the cycle, the value stored depends on
the value loaded.
This dependency might be a data dependency
(see Listing~\ref{lst:OOTA Cycle})
or it might be a control dependency
(see Listing~\ref{lst:OOTA Cycle Based on Conditionals}).\footnote{
	Sufficiently motivated readers should have no problem constructing
	examples involving address dependencies.}
There must be at least two distinct values such that if a given load
returns a given value, that value forces the cycle to execute so as to
cause the corresponding store's value to be equal to that given value, and
for that store to occur early enough for the load to return that value.

@@

This atemporality is also why cycles are not considered to have
OOTA cycles when at least one link from one thread to the next
involves both threads storing to the same object.
Similarly, cycles having at least link where one thread loads from an
object and the next thread stores to that object are also not considered
to be OOTA cycles.

@@

\section{Wording}
\label{sec:Wording}

Add the following note in \co{[atomics.order]} after 33.5.4p9:

\begin{quote}
	[ \emph{Note:} Implementations that respect the following
	constraints are incapable of computing out-of-thin-air value
	that depend circularly on their own computation:

	\begin{itemize}
	\item	Correctly implement the C++ abstract machine, and
	\item	Run on correctly implemented hardware
		(specifically, hardware that forbids non-speculative
		loads from returning values from speculative stores
		and that forbids exposing values from speculative
		stores to other threads), and
	\item	Reside in a universe having a finite speed of light
		and atomic matter.
	\end{itemize}

	All three constraints are subject to the as-if rule.
	For example, those creating an alternate universe in which
	C++ implementations are to run correctly should ensure
	that their universe behaves as if the speed of light was
	finite and the nature of matter was atomic.
	--- \emph{end note} ]
\end{quote}

@@

We are grateful to David Goldblatt for his careful review of an early
draft of this paper and to John Wikerson for asking Paul for a rant and
taking the proffered rant seriously.
We also owe David Goldblatt a debt of gratitude for his having asked
an insightful question at the right time.
Nonetheless, all errors and omissions in this paper are the sole property
of the authors.

% To do:
% Luc Maranget: Memory models.
% Richard Gristenthwaite: ARM hardware.
% Mark Rutland: ARM memory ordering.
% Will Deacon: ARM hardware, C++, and OOTA.
% Greg Kroah-Hartman and Linus Torvalds: FYI, trouble being caused.
% Christoph Hellwig: Why slow on BPF memory model.
% Miguel Ojeda: LKMM and Rust.
% Alice Ryhl: LKMM and Rust.
% Ralf Jung: Rust.
% Catalin Marinas: ARM and formal methods (qspinlock).
% Segher Boessenkool:  GCC.
% Dan Lustig: NVIDIA memory ordering.
% Andrea Parri: Memory models, RISC-V hardware.
% Jonas Oberhauser: Memory models.
% Hernan Ponce de Leon: Memory models.
% Palmer Dabbelt: RISC-V hardware.
% Ali Sezgin: P0422R0.
% Tony Tye: P0422R0.
% Other memory-model maintainers, including Peter Zijlstra.
% Derek Williams: PowerPC hardware.
% Hans Boehm: Skeptic, memory models, OOTA.  Suggestions for others?
% Olivier Giroux: Skeptic, memory models, NVIDIA.
% Luke Geeson: Whatever...  Review.
% Ori Lahav: Prior OOTA.
% Viktor Vafeiadis:  Prior OOTA.
% Brian Demsky: Prior OOTA.

% @@@ Inter-CPU speculation if inter-CPU squashing of speculation.

% Done:
% David Goldblatt
% Michael Wong
% Maged Michael
% John Wickerson: Thank you.  Peter Sewell?  Mark Batty? Other TFWM authors?
% Akira Yokosawa: Professional courtesy, perfbook memory models.
% Dan Kelley, Alexei Starovoitov, Mykola Lysenko: FYI.  OOTA dropped
% Alan Jeffrey: Fixed-point insight P0422R0.  Reached out via LinkedIn.
% Jade Alglave: C++ herd model?  Memory models and variety of hardware.
% Alan Stern: Memory models, mathematical logic, and variety of hardware.

% \appendix

% \section{Demonstration of OOTA-Cycle Definition}
% \label{app:Demonstration of OOTA-Cycle Definition}
